### 2018-03-13 submission/manuscript.tex
add pat's result edits

\textbf{Background.} An increasing body of literature suggests that both
individual and collections of bacteria are associated with the
progression of colorectal cancer. As the number of studies investigating
these associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. For our
meta-analysis, we analyzed previously published 16S rRNA gene sequencing
data collected from feces (1737 individuals from 8 studies) and colon
tissue (492 total samples from 350 individuals from 7 studies).

\textbf{Results.} We quantified the odds ratios for individual bacterial
taxa that were associated with an individual having tumors relative to a
normal colon. Among the stool samples, there were no taxa that had a
significant odds ratio associated with adenoma and there were 8 taxa
with significant odds ratios (ORs) associated with carcinoma. Similarly,
among the tissue samples, there were no taxa that had a significant odds
ratio associated with adenoma and there were 3 taxa with significant
odds ratios associated with carcinoma. Among the significant odds
ratios, the association between individual taxa and tumor diagnosis was
equal or below 7.11. Because individual taxa had limited association
with tumor diagnosis, we trained Random Forest classification models
using only the taxa that had significant ORs, using the entire
collection of taxa found in each study, and using operational taxonomic
units defined based on a 97\% similarity threshold. All training
approaches yielded similar classification success as measured using the
Area Under the Curve. The ability to correctly classify individuals with
adenomas was poor and the ability to classify individuals with
carcinomas was considerably better using sequences from stool or tissue.

@@

\textbf{Importance.} This meta-analysis confirms previous results
indicating that individuals with adenomas cannot be readily classified
based on their bacterial community, but that those with carcinomas can.
Regardless of the dataset, we found a subset of the fecal community that
was associated with carcinomas was as predictive as the full community.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been hypothesized to have a role in disease
progression (1, 2). Numerous studies using murine models of CRC have
shown the importance of both individual microbes (3--7) and the overall
community (8--10) in tumorigenesis. Numerous case-control studies have
characterized the microbiota of individuals with colonic adenomas and
carcinomas in an attempt to identify biomarkers of disease progression
(6, 11--17). Because current CRC screening recommendations are poorly
adhered to due to socioeconomic status, test invasiveness, and frequency
of tests, development and validation of microbiome-associated biomarkers
for CRC progression could further attempts to develop non-invasive
diagnostics (18).

Recently, there has been an intense focus on identifying
microbiota-based biomarkers yielding a seemingly endless number of
candidate taxa. Some studies point towards mouth-associated genera such
as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
are enriched in people with carcinomas (6, 11--17). Other studies have
identified members of \emph{Akkermansia}, \emph{Bacteroides},
\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
also associated with carciomas (13--15). Additionally, \emph{Roseburia}
has been found in some studies to be more abundant in people with tumors
but in other studies it has been found to be either less abundant or no
different than what is found in subjects with normal colons (14, 17, 19,
20). There are strong results from tissue culture and murine models that
\emph{Fusobacterium nucleatum}, pks-positive strains of
\emph{Escherichia coli}, \emph{Streptococcus gallolyticus}, and an
entertoxin-producing strain of \emph{Bacteroides fragilis} are important
in the pathogenesis of CRC (5, 14, 21--24). These results point to a
causative role for the microbiota in CRC pathogenesis as well as their
potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with
carcinomas but there is a greater clinical need to identify biomarkers
associated with adenomas. Studies focusing on broad scale community
metrics have found that measures such as the total number of Operational
Taxonomic Units (OTUs) are decreased in those with adenomas versus
controls (25). Other studies have identified \emph{Acidovorax},
\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
with adenomas (25--27). There are few genera that are enriched in
patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found
that 16S rRNA gene sequences from members of \emph{Akkermansia},
\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
the presence of carcinomas (28). Contrary to previous studies they found
sequences similar to members of \emph{Lactobacillus} and
\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
relative to those with normal colons (12, 15, 16). In addition, they
found 16S rRNA gene sequences from members of \emph{Haemophilus},
\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
enriched in patients with adenoma and \emph{Pantoea} were enriched in
patients with carcinomas. Although this meta-analysis was helpful for
distilling a large number of possible biomarkers, the aggregate number
of samples included in the analysis (n = 509) was smaller than several
larger case-control studies that have since been published (12, 27)

Here we provide an updated meta-analysis using 16S rRNA gene sequence
data from both feces (n = 1737) and colon tissue (492 samples from 350
individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
\& 2{]}. We expand both the breadth and scope of the previous
meta-analysis to investigate whether biomarkers describing the bacterial
community or specific members of the community can more accurately
classify patients as having adenoma or carcinoma. Our results suggest
that the bacterial community changes as disease severity worsens and
that a subset of the microbial community can be used to diagnose the
presence of carcinoma.

@@

\textbf{\emph{Lower bacterial diversity is associated with increased
odds ratio (OR) of tumors.}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was no significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65)(95\% Confidence Interval), P-value =
0.035) and carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure
2B{]}. The ORs for diversity were only significantly greater than 1.0
for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069), but not for
adenoma (P-value = 0.11) {[}Figure 2C{]}. Although these OR are
significantly greater than 1.0, it is doubtful that these are clinically
meaningful values.

Similar to our analysis of sequences obtained from stool samples, we
repeated the analysis using sequences obtained from colon tissue. There
were no significant changes in richness, evenness, or diversity as
disease severity progressed from control to adenoma to carcinoma
(P-value \textgreater{} 0.05). We next analyzed the OR, for matched
(i.e.~where unaffected tissue and tumors were obtained from the same
individual) and unmatched (i.e.~where unaffected tissue and tumor tissue
were not obtained from the same individual) tissue samples. The ORs for
adenoma and carcinoma by any measure were not significantly different
from 1.0 (P-value \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
is likely due to the combination of a small effect size, as suggested
from the results using stool, and the relatively small number of studies
and size of studies used in the analysis.

\textbf{\emph{Disease progression is associated with community-wide
changes in composition and abundance.}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there are consistent and significant community-wide
changes in the fecal community structure of subjects with carcinomas.
However, the signal observed in subjects with adenomas or when using
tissue samples was not as consistent. This is likely due to a smaller
effect size or the relatively small sample sizes among the studies that
characterized the tissue microbiota.

\textbf{\emph{Individual taxa are associated with significant ORs for
carcinomas.}} Next we identified those taxa that were associated with
ORs that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
0.83)). Among the data collected from tissue samples, only unmatched
carcinoma samples had taxa with a significant OR. Those included
\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
Mouth-associated genera were not significantly associated with an
increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
due to the small number of studies and considerable variation in the
data, the Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}.
It is interesting to note that \emph{Ruminococcus} and members of
\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
tissue had ORs that were significantly less than 1.0, which suggests
that these populations are protective against the development of
carcinomas. Overall, there was no overlap in the taxa with significant
OR between stool and tissue samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma.}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. OR values were
caluclated based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study. To measure the ability of these taxa to
classify individuals we instead generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included \emph{Clostridium XI},
\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
AUC values less than 0.588. Likewise, in unmatched tissue samples the 3
taxa with significant OR taxa were marginally better than one would
expect by chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea}
was the best predictor of carcinomas and its median AUC was only 0.62.
These results suggest that although these taxa are associated with a
decreased or increased OR for the presences of carcinomas, individually,
they do a poor job of classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa.}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). Within the
significant OR taxa only models the most important taxa across study for
stool were \emph{Ruminococcus} and \emph{Clostridium XI} based on mean
decrease in accuracy {[}Figure 4A{]}. Similarly, using data from the
unmatched tissue samples, the combined model had an AUC of 0.77, which
was significantly higher than the AUC values for \emph{Blautia} and
\emph{Weissella} (P-value \textless{} 0.037). For the significant OR
taxa unmatched tissue models both \emph{Dorea} and \emph{Blautia} were
the most important based on mean decrease in accuracy {[}Figure 4B{]}.
Pooling the information from the taxa with significant ORs results in a
model that outperforms classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs.}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC = 0.549
(0.367 - 0.971)(min-max)) {[}Figure 5A \& S2A{]}. In contrast, the
models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 5B \& S2B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of 14.8\%) and AUC for the Flemer study decreased by 0.54\%).
The overall improvement in performance was statistically significant
(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). For these
full taxa models using stool, \emph{Bacteroides} and
\emph{Lachnospiraceae} were the most common taxa in the top 10\% mean
decrease in accuracy (MDA) across studies {[}Figure S3{]}. Similarly,
using the data from unmatched tissue samples we found that the AUC of
studies increased by an average of 19.11\% when we used all of the taxa
rather than the 3 taxa with significant ORs (one-tailed paired T-test;
P-value = 0.03). For the full taxa models using unmatched tissue,
\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies {[}Figure S4{]}. Although the
significant OR taxa models can classify individuals with and without
carcinoma tumors, they are still missing taxa from the full community
models that can increase the model accuracy.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs.}} The previous models were based on relative
abundance data where sequences were classified to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)), but did
a good job of differentiating between subjects with normal colons and
those with carcinomas (median AUC: 0.71 (0.5- 0.9)). The OTU-based
models performed similarly to those constructed using the taxa with
significant ORs (one-tailed paired T-test; P-value = 0.966) and those
using all taxa (one-tailed paired T-test; P-value = 0.146) {[}Figure
4{]}. Among the OTUs that had the highest mean decrease in accuracy for
the OTU-based models, we found that OTUs that affiliated with all of the
8 taxa that had a significant OR were within the top 10\% for at least
one study. This result was surprising as it indicated that a finer scale
classification of the sequences and thus a larger number of features to
select from, did not yield improved classification of the subjects.

>>>>

An increasing body of literature suggests that both individual and
collections of bacteria are associated with the progression of
colorectal cancer. As the number of studies investigating these
associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. We analyzed
previously published 16S rRNA gene sequencing data collected from feces
and colon tissue.

We quantified the odds ratios (ORs) for individual bacterial taxa that
were associated with an individual having tumors relative to a normal
colon. Among the fecal samples, there were no taxa that had a
significant ORs associated with adenoma and there were 8 taxa with
significant ORs associated with carcinoma. Similarly, among the tissue
samples, there were no taxa that had a significant OR associated with
adenoma and there were 3 taxa with significant ORs associated with
carcinoma. Among the significant ORs, the association between individual
taxa and tumor diagnosis was equal or below 7.11. Because individual
taxa had limited association with tumor diagnosis, we trained Random
Forest classification models using only the taxa that had significant
ORs, using the entire collection of taxa found in each study, and using
operational taxonomic units defined based on a 97\% similarity
threshold. All training approaches yielded similar classification
success as measured using the Area Under the Curve. The ability to
correctly classify individuals with adenomas was poor and the ability to
classify individuals with carcinomas was considerably better using
sequences from fecal or tissue.

@@

\subsection{Importance}\label{importance}

Colorectal cancer is a significant and growing health problem in which
animal models and epidemiological data suggest that the colonic
microbiota have a role in tumorigenesis. These observations indicate
that the colonic microbiota is a reservoir of biomarkers that may
improve our ability to detect colonic tumors using non-invasive
approaches. This meta-analysis identifies and validates a set of 8
bacterial taxa that can be used within a Random Forest modeling
framework to differentiate individuals as having normal colons or
carcinomas. When models trained using one dataset were tested on other
datasets, the models performed well. These results lend support to the
use of fecal biomarkers for the detection of tumors. Furthermore, these
biomarkers are plausible candidates for further mechanistic studies into
the role of the gut microbiota in tumorigenesis.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been hypothesized to have a role in disease
progression (1, 2). Numerous studies using murine models of CRC have
shown the importance of both individual microbes (3--7) and the overall
community (8--10) in tumorigenesis. Numerous case-control studies have
characterized the microbiota of individuals with colonic adenomas and
carcinomas in an attempt to identify biomarkers of disease progression
(6, 11--17). Because current CRC screening recommendations are poorly
adhered to due to person's socioeconomic status, test invasiveness, and
frequency of tests, development and validation of microbiota-associated
biomarkers for CRC progression could further attempts to develop
non-invasive diagnostics (18).

Recently, there has been an intense focus on identifying
microbiota-based biomarkers yielding a seemingly endless number of
candidate taxa. Some studies point towards mouth-associated genera such
as \emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas},
and \emph{Porphyromonas} that are enriched in people with carcinomas (6,
11--17). Other studies have identified members of \emph{Akkermansia},
\emph{Bacteroides}, \emph{Enterococcus}, \emph{Escherichia},
\emph{Klebsiella}, \emph{Mogibacterium}, \emph{Streptococcus}, and
\emph{Providencia} (13--15). Additionally, \emph{Roseburia} has been
found in some studies to be more abundant in people with tumors but in
other studies it has been found to be less abundant than what is found
in subjects with normal colons (14, 17, 19, 20). There is support from
mechanistic studies using tissue culture and murine models that
\emph{Fusobacterium nucleatum}, pks-positive strains of
\emph{Escherichia coli}, \emph{Streptococcus gallolyticus}, and an
entertoxin-producing strain of \emph{Bacteroides fragilis} are important
in tumorigenesis (5, 14, 21--24). These results point to a causative
role for the microbiota in tumorigenesis as well as their potential as
diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with
carcinomas but there is a clinical need to identify biomarkers
associated with adenomas to facilitate early detection of the tumors.
Studies focusing on broad scale community metrics have found that
measures such as the total number of taxa (i.e.~richness) are lower in
those with adenomas versus controls (25). Other studies have identified
\emph{Acidovorax}, \emph{Bilophila}, \emph{Cloacibacterium},
\emph{Desulfovibrio}, \emph{Helicobacter}, \emph{Lactobacillus},
\emph{Lactococcus}, \emph{Mogibacterium}, and \emph{Pseudomonas} to be
enriched in those with adenomas (25--27). The ability to classify
individuals as having normal colons or adenomas based solely on the taxa
within fecal samples has been limited. However, when 16S rRNA gene
sequence data was combined with the results of a fecal immunochemical
test (FIT), the ability to diagnose individuals with adenomas was
improved relative to using the FIT results alone (12).

A recent meta-analysis found that 16S rRNA gene sequences from members
of \emph{Akkermansia}, \emph{Fusobacterium}, and \emph{Parvimonas} were
fecal biomarkers for the presence of carcinomas (28). Contrary to
previous studies they found sequences similar to members of
\emph{Lactobacillus} and \emph{Ruminococcus} to be enriched in patients
with adenoma or carcinoma relative to those with normal colons (12, 15,
16). In addition, they found that 16S rRNA gene sequences from members
of \emph{Haemophilus}, \emph{Methanosphaera}, \emph{Prevotella}, and
\emph{Succinovibrio} were enriched in patients with adenomas and
\emph{Pantoea} were enriched in patients with carcinomas. Although this
meta-analysis was helpful for distilling a large number of possible
biomarkers, the aggregate number of samples included in the analysis
(n=509) was smaller than several larger case-control studies that have
since been published (12, 27)

Here we provide an updated meta-analysis using 16S rRNA gene sequence
data from both feces (n=1737) and colon tissue (492 samples from 350
individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
\& 2{]}. We expand both the breadth and scope of the previous
meta-analysis to investigate whether biomarkers describing the bacterial
community or specific members of the community can more accurately
classify patients as having adenoma or carcinoma. Our results suggest
that the bacterial community changes as disease severity worsens and
that a subset of the microbial community can be used to diagnose the
presence of carcinoma.

@@

\textbf{\emph{Lower bacterial diversity is associated with higher odds
ratio (OR) of tumors.}} We first assessed whether variation in broad
community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity of the communities were associated with disease stage after
controlling for study and variable region differences. In fecal samples,
both evenness and diversity were significantly lower in successive
disease severity categories (evenness P-value=0.025 and diversity
P-value=0.043) {[}Figure 1{]}; there was no significant difference for
richness (P-value=0.21). We next tested whether the lower value of these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value=0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR=1.3 (95\% Confidence Interval: 1.02 - 1.65),
P-value=0.035) and carcinoma (OR=1.66 (1.2 - 2.3), P-value=0.0021)
{[}Figure 2B{]}. The ORs for diversity were only significantly greater
than 1.0 for carcinoma (OR=1.61 (1.14 - 2.28), P-value=0.0069), but not
for adenoma (P-value=0.11) {[}Figure 2C{]}. Although these ORs are
significantly greater than 1.0, it is doubtful that they are clinically
meaningful.

Similar to our analysis of sequences obtained from fecal samples, we
repeated the analysis using sequences obtained from colon tissue. There
were no significant differences in richness, evenness, or diversity as
disease severity progressed from control to adenoma to carcinoma
(P-values \textgreater{} 0.05). We next analyzed the ORs, for matched
(i.e.~where unaffected tissue and tumors were obtained from the same
individual) and unmatched (i.e.~where unaffected tissue and tumor tissue
were not obtained from the same individual) tissue samples. The ORs for
adenoma and carcinoma were not significantly different from 1.0 for any
measure (P-values \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
is likely due to the combination of a small effect size and the
relatively small number of studies and size of studies used in the
analysis.

\textbf{\emph{Disease progression is associated with changes in
community structure.}} Based on the differences in evenness and
diversity, we next asked whether there were community-wide differences
in the structure of the communities associated with different disease
stages. We identified significant bacterial community differences in the
feces of patients with adenomas relative to those with normal colons in
1 of 4 studies and in patients with carcinomas relative to those with
normal colons in 6 of 7 studies (PERMANOVA; P-value \textless{} 0.05)
{[}Table S2{]}. Similar to the analyses using fecal samples, there were
significant differences in the bacterial community structure of subjects
with normal colons and those with adenomas (1 of 2 studies) and
carcinomas (1 of 3 studies) {[}Table S2{]}. For studies that used
matched samples, we did not observe any differences in bacterial
community structures {[}Table S2{]}. Combined, these results indicate
that there were consistent and significant community-wide changes in the
fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual taxa are associated with significant ORs for
carcinomas.}} Next we identified those taxa that had ORs that were
significantly associated with having a normal colon or the presence of
adenomas or carcinomas. No taxa had a significant OR for the presence of
adenomas when we used data collected from fecal or tissue samples (Table
S3 \& S4). In contrast, 8 taxa had significant ORs for the presence of
carcinomas using data from fecal samples. Of these, 4 are commonly
associated with the oral cavity: \emph{Fusobacterium} (OR=2.74 (1.95 -
3.85)), \emph{Parvimonas} (OR=3.07 (2.11 - 4.46)), \emph{Porphyromonas}
(OR=3.2 (2.26 - 4.54)), and \emph{Peptostreptococcus} (OR=7.11 (3.84 -
13.17)) {[}Table S3{]}. The other 4 were \emph{Clostridium XI} (OR=0.65
(0.49 - 0.86)), \emph{Enterobacteriaceae} (OR=1.79 (1.33 - 2.41)),
\emph{Escherichia} (OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus}
(OR=0.63 (0.48 - 0.83)). Among the data collected from tissue samples,
only unmatched carcinoma samples had taxa with a significant OR. Those
included \emph{Dorea} (OR=0.35 (0.22 - 0.55)), \emph{Blautia} (OR=0.47
(0.3 - 0.73)), and \emph{Weissella} (OR=5.15 (2.02 - 13.14)).
Mouth-associated genera were not significantly associated with a higher
OR for carcinoma in tissue samples {[}Table S4{]}. For example,
\emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however, due to
the small number of studies and considerable variation in the data, the
Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}. It is
interesting to note that \emph{Ruminococcus} and members of
\emph{Clostridium XI} in fecal samples and \emph{Dorea} and
\emph{Blautia} in tissue had ORs that were significantly less than 1.0,
which suggests that these populations are protective against the
development of carcinomas. Overall, there was no overlap in the taxa
with significant OR between fecal and tissue samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma.}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. OR values were
caluclated based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study. To measure the ability of these taxa to
classify individuals we instead generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for classifying individuals by their
disease status. Using data from fecal samples, the 8 taxa did no better
at classifying the subjects than one would expect by chance
(i.e.~AUC=0.50) {[}Figure 3A{]}. The taxa that performed the best
included \emph{Clostridium XI}, \emph{Ruminococcus}, and
\emph{Escherichia}. However, these had median AUC values less than 0.588
ndicating their limited value as biomarkers when used individually.
Likewise, in unmatched tissue samples the 3 taxa with significant OR
taxa had AUC values that were marginally better than one would expect by
chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea} was the
best predictor of carcinomas and its median AUC was only 0.62. These
results suggest that although these taxa are associated with a
significant OR for the presences of carcinomas, they do a poor job of
classifying a subject's disease status when used individually.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa.}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from fecal samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). When this
approach was used to train models using data from each study, the most
important taxa were \emph{Ruminococcus} and \emph{Clostridium XI}
{[}Figure 4A{]}. Similarly, using data from the unmatched tissue
samples, the combined model had an AUC of 0.77, which was significantly
higher than the AUC values for classifying based on the relative
abundances of \emph{Blautia} and \emph{Weissella} individually (P-value
\textless{} 0.037). Both \emph{Dorea} and \emph{Blautia} were the most
important taxa in the tissue-based models {[}Figure 4B{]}. Pooling the
information from the taxa with significant ORs resulted in models that
outperformed classifications made using the same taxa individually.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community is better than that of models based on taxa with
significant ORs.}} Next, we asked whether a Random Forest classification
model built using all of the taxa found in the communities would
outperform the models generated using those taxa with a significant OR.
Similar to our inability to identify taxa associated with a significant
OR for the presence of adenomas, the median AUCs to classify subjects as
having normal colons or having adenomas using data from fecal or tissue
samples were only marginally better than 0.5 for any study (median
AUC=0.549 (range: 0.367 - 0.971)) {[}Figure 5A \& S2A{]}. In contrast,
the models for classifying subjects as having normal colons or having
carcinomas using data from fecal or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 5B \& S2B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from fecal samples, we found that the AUC for 6 of 7 studies were an
average of 14.8\% higher and AUC for the Flemer study was 0.54\% lower
when using the relative abundance data from all taxa relative to using
the relative abundance of only the taxa with significant ORs. The
overall improvement in performance was statistically significant
(mean=12.61\%, one-tailed paired T-test; P-value=0.005). Among the
models trained using data from fecal samples, \emph{Bacteroides} and
\emph{Lachnospiraceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies {[}Figure S3{]}. Using data from
unmatched tissue samples to train classification models, we found that
the AUC of studies was an average 19.11\% higher when we used all of the
taxa rather than the 3 taxa with significant ORs (one-tailed paired
T-test; P-value=0.03). For the models trained using data from unmatched
tissue samples, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies {[}Figure S4{]}. Although the models
trained using those taxa with a significant OR perform well for
classifying individuals with and without carcinomas, models trained
using data from the full community perform better.

\textbf{\emph{Performance of models based on OTU relative abundances are
not significantly better than those based on taxa with significant
ORs.}} The previous models were based on relative abundance data where
sequences were classified to coarse taxonomic assignments
(i.e.~typically genus or family level). To determine whether model
performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)). However,
they did a good job of differentiating between subjects with normal
colons and those with carcinomas (median AUC: 0.71 (0.50- 0.904)). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value = 0.966) and
those using all taxa (one-tailed paired T-test; P-value = 0.146)
{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
accuracy for the OTU-based models, we found that OTUs that affiliated
with all of the 8 taxa that had a significant OR were within the top
10\% for at least one study. This result was surprising as it indicated
that a finer scale classification of the sequences and thus a larger
number of features to select from, did not yield improved classification
of the subjects.
